---
title: "Assignment 1"
author: "Yuki Joyama (yj2803)"
output: 
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{amsmath}  # Include amsmath package
  - \newcommand{\indep}{\perp\!\!\!\perp}  # Define the conditional independence symbol
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)

library(tidyverse)
library(ggplot2)
library(dagitty)

# setup plot theme
theme_set(
  theme_bw() +
    theme(legend.position = "top")
  )
```

# Problem 1
(a) $X_2 \indep X_3, X_7, X_8, X_{10}$  
$X_3 \indep X_2|X_1$  
$X_4 \indep X_1, X_7, X_8, X_{10}|X_2, X_3$  
$X_5 \indep X_1, X_2, X_3, X_7, X_8, X_9, X_{10}|X_4$  
$X_6 \indep X_1, X_2, X_3, X_4, X_7, X_8, X_9, X_{10}|X_5$  
$X_7 \indep X_1, X_2, X_4, X_5, X_6|X_3$  
$X_8 \indep X_1, X_2, X_4, X_5, X_6|X_3, X_7$  
$X_9 \indep X_1, X_2, X_3, X_5, X_6, X_7|X_4, X_8, X_{10}$  
$X_{10} \indep X_1, X_2, X_3, X_4, X_5, X_6, X_7|X_8$  

(b) $X_2 \perp_dX_9|X_4$: False   
$X_7 \perp_dX_5|\{X_3, X_8\}$: True  
$\{X_2, X_4\} \perp_dX_7|\{X_6, X_9, X_{10}\}$: False
```{r echo=F, results=F}
gb <- dagitty('dag {
    X1 [pos="0,0"]
    X2 [pos="-1,1"]
    X3 [pos="2,1"]
    X4 [pos="0,2"]
    X5 [pos="0,3"]
    X6 [pos="-1,3"]
    X7 [pos="3,1"]
    X8 [pos="2.5,2"]
    X9 [pos="2.5,3"]
    X10 [pos="4,2"]
    
    X1 -> X2 -> X4 -> X5 -> X6
    X1 -> X3 -> X4 -> X9
    X3 -> X7 -> X8 -> X10 -> X9
    X8 -> X9
    X3 -> X8
}')

dseparated(gb, "X2", "X9", c("X4"))
dseparated(gb, "X7", "X5", c("X3", "X8"))
dseparated(gb, c("X2", "X4"), "X7", c("X6", "X9", "X10"))
```

# Problem 2


# Problem 3
Agreements: $D \indep \{A,C\}|B$  
Disagreements: $A \indep C|\phi$ (a), $C \indep \{A,D\}|B$ (b)

# Problem 4
(a) Unshielded colliders in Figure 3: $C\rightarrow E \leftarrow A$, $C\rightarrow B \leftarrow A$, $D\rightarrow E \leftarrow B$  
No DAGs in the option matches these three structures. 

(b) An unshielded collider in a DAG will violate the Markov equivalence with respect to the chain DAG. Taken this into account, we can consider two cases that will be Markov equivalent to the chain DAG:  
1.Flip all the arrows to left  
2.For $X_i (i=2, 3, ..., p-1)$, flip all the arrows to left before $X_i$  
There are p-2 possible $X_i$s that can be the pivot of the arrows. Therefore, $p-2+1=p-1$ DAGs are Markov equivalent to the chain DAG. 

# Problem 5
(a)
```{r results=T, echo=F}
# problem 5
library(dagitty)

# construct fig 5 DAG
g <- dagitty('dag {
    D [pos="0,0"]
    E [pos="1,0"]
    C [pos="1,-1"]
    A [pos="2,0"]
    B [pos="3,0"]
    F [pos="4,0"]
    G [pos="4,-1"]
    H [pos="5,-1"]
    
    D -> E -> A <- B <- G -> H
    C -> E -> F -> G
    C -> H
    C -> B
    C -> F -> H
}')

# a: path from C to H
paths(g, "C", "H")
```

(b)
```{r message=T, echo=F}
# b: d-separation between E and G given A and B
if(dseparated(g, "E", "G", c("A", "B"))){
  message("E"," and ", "G"," are d-separated given A and B.")
} else {
  message("E"," and ", "G", " are not d-separated given A and B.")
}
```

(c)
```{r results=T, echo=F}
# c: list the conditional independencies relationships implied by the model
impliedConditionalIndependencies(g)
```

With the option “type = all.pairs”:
```{r results=T, echo=F}
impliedConditionalIndependencies(g, type = "all.pairs")
```
According to the documentation, the default for this function is type = "missing.edge". This returns a list of conditional independencies with minimal testable implication per missing edge while type = "all.pairs" returns all implied conditional independencies between two variables. This is why the first one is shorter than the second one.

(d) The summary of the simulated data (N=10000) is shown below.
```{r results=T, echo=F}
set.seed(2024)
# d: simulate data from this DAG, which associates the DAG with a linear structural equation model
# path coefficient (-0.7, 0.7), sample size = 10000
sim <- simulateSEM(
  g,
  b.default = NULL,
  b.lower = -0.7,
  b.upper = 0.7,
  N = 10000
) 

summary(sim)
```

Markov blanket for vertex $B$ are:
```{r results=T, echo=F}
markovBlanket(g, 'B')
```
Let's check the linear regression of $B \sim Mb(B, \mathcal{G})$ + remaining covariates using this simulated data.
```{r results=T, echo=F}
# construct a linear model
lm_b = lm(B ~ A + C + D + E + F + G + H, sim)
summary(lm_b)
```
The coefficients for variables outside of the Markov blanket ($D, F, H$) have p-values greater than 0.05, indicating their independence from $B$. Meanwhile, the coefficients for variables within the Markov blanket have p-values less than 0.05, confirming that the Markov blanket property holds for vertex $B$.

# Code
```{r echo=T, results=F, message=F}
# problem 5
library(dagitty)

# construct fig 5 DAG
g <- dagitty('dag {
    D [pos="0,0"]
    E [pos="1,0"]
    C [pos="1,-1"]
    A [pos="2,0"]
    B [pos="3,0"]
    F [pos="4,0"]
    G [pos="4,-1"]
    H [pos="5,-1"]
    
    D -> E -> A <- B <- G -> H
    C -> E -> F -> G
    C -> H
    C -> B
    C -> F -> H
}')

# a: path from C to H
paths(g, "C", "H")

# b: d-separation between E and G given A and B
if(dseparated(g, "E", "G", c("A", "B"))){
  message("E"," and ", "G"," are d-separated given A and B.")
} else {
  message("E"," and ", "G", " are not d-separated given A and B.")
}

# c: list the conditional independencies relationships implied by the model
impliedConditionalIndependencies(g)
impliedConditionalIndependencies(g, type = "all.pairs")

set.seed(2024)
# d: simulate data from this DAG, which associates the DAG with a linear structural equation model
# path coefficient (-0.7, 0.7), sample size = 10000
sim <- simulateSEM(
  g,
  b.default = NULL,
  b.lower = -0.7,
  b.upper = 0.7,
  N = 10000
)

summary(sim)

markovBlanket(g, 'B')

# construct a linear model
lm_b = lm(B ~ A + C + D + E + F + G + H, sim)
summary(lm_b)
```

